---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# SimRVPedigree
Simulate Pedigrees Ascertained for Rare Disease

# Under Construction

Please note: this version of `SimRVPedigree` may be under construction.  For the latest stable version of `SimRVPedigree` please visit <https://CRAN.R-project.org/package=SimRVPedigree>.

## Installation

You can install SimRVPedigree from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("cnieuwoudt/SimRVPedigree")
```

## Overview
`SimRVPedigree` provides routines to simulate and manipulate pedigrees ascertained to contain multiple family members affected by a rare disease.

To simulate pedigrees or to simulate life events for an individual we provide:
* `sim_RVped()` to simulate pedigrees ascertained for multiple disease-affected relatives,
* `sim_ped()` to simulate pedigrees, and
* `sim_life()` to simulate life events until a specified stop-year.

To manipulate pedigrees we provide:
* `plot.ped()` to plot pedigrees,
* `summary.ped()` to obtain summary information for a pedigree or a collection of pedigrees,
* `censor_ped()` to censor a pedigree after a specified year, and
* `reassign_gen()` to reassign generation number based on the most recent common ancestor of all disease-affected relatives.

## Example

```{r example, fig.height=6, fig.width=6}
library(SimRVPedigree)

#Load example age-specific hazard rates data.
data(AgeSpecific_Hazards)

#Create an object of class hazard.
HRates <- hazard(hazardDF = AgeSpecific_Hazards)

#Simulate a pedigree ascertained for multiple disease-affected relatives
set.seed(5444)
ex_ped <- sim_RVped(hazard_rates = HRates,
                    GRR = 20,
                    RVfounder = TRUE,
                    FamID = 1,
                    founder_byears = c(1900, 1950),
                    ascertain_span = c(1995, 2015),
                    num_affected = 2,
                    stop_year = 2017,
                    recall_probs = c(1, 1, 0.25, 0))

# Plot the ascertained pedigree
plot(ex_ped[[2]], ref_year = 2017, cex = 0.75)

# Obtain summary information for the ascertained pedigree.
summary(ex_ped[[2]])
```
